detect patches

### 2024-02-13 scripts/classify_patches.Rmd
text

```{r}
library(terra)
library(tidyverse)
library(sf)
library(landscapemetrics)
```

@@

r1 <- rast(f[1])

patchTaggR <- function(x){

  r <- terra::rast(x)
  r_taxon <- gsub('_', ' ', gsub('-patchIDS[.]tif', '', basename(x)))
  occs <- dplyr::filter(recs, taxon == r_taxon) |>
    dplyr::mutate(ID = dplyr::row_number()) |>
    terra::vect()

  occ_pts <- terra::extract(r, occs, bind = TRUE) |>
    sf::st_as_sf()

  # if pts were NA, buffer to 90m, many occupied areas were 'cut' away by the 
  # hydrologic mapping. 
  occ_poly <- dplyr::filter(occ_pts, is.na(lyr.1)) |>
    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.

  occ_poly <- terra::extract(r, occ_poly, weights = TRUE) %>% 
    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
    tidyr::drop_na(lyr.1) |> # these values actually NA. 
    dplyr::slice_max(weight, with_ties = F) |>
    dplyr::left_join(filter(occ_pts) |>
                       select(-lyr.1), by = join_by('ID')) |>
    dplyr::select(-weight)

  occ_pops <- dplyr::bind_rows( # a data frame of all raster patch ID's with populations known from within
    # the last few decades in them. 
    tidyr::drop_na(occ_pts, lyr.1), occ_poly) |>
    dplyr::rename(PatchID = lyr.1)

  # write out summary values of the patch matching
  data.frame(
    noPres = , 
    PresTopSuitability =, # the number of presences located in top suitability patches
    UniquePatchID = , # the number of uniquelly identified patches
    NoPatchesTopSuitability =  # the total number of most suitable patches from the RASTER
  )

  occ_pops <- dplyr::distinct(occ_pops, PatchID, .keep_all = TRUE) |> # this is multiple pts per patch - not necessary for us. 
    sf::st_drop_geometry() |>
    dplyr::select(date, PatchID) |>
    dplyr::arrange(PatchID) |>
    write.csv(row.names = F,
            file.path('../results/occupiedPatches-topSuitability', paste0(r_taxon, '.csv')))

  # for points which were not patched to a most suitable patch, put them into a hydrologic basin

@@

patchTaggR(f[1])
```

>>>>

```{r}
library(terra)
library(tidyverse)
library(sf)
library(landscapemetrics)
source('functions.R')
```

@@

gd <- '/media/steppe/hdd/Geospatial_data'
hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu12R-line.tif') )
hu12 <- rast('../../Geospatial_data/WBD_rast/hu12R.tif')

lapply(f, patchTaggR)
```

Combine all populated patches per species, something was amiss with identifying patches in marginal habitat, we will thin them down here.

f <- gsub('[.]tif$', '', list.files('../results/patches'))

combine_pop_patches <- function(x){

  pIDs <- read.csv(paste0('../results/occupiedPatches/', x, '-topSuitability.csv'))
  top_patches <- terra::rast(paste0('../results/fine_patches/', x, '-patchIDS.tif'))
  highest_no <- minmax(top_patches)[2]
  top_patches <- terra::subst(top_patches, pIDs$PatchID,  pIDs$PatchID, others = NA)
  names(top_patches)<- 'patches'
  pts <- st_read(paste0('../data/raw/occurrence/combined_records/', x, '.shp'), quiet = TRUE) |>
    dplyr::filter(Occurrence == 1)
  marg_patch <- terra::rast(paste0('../results/marginal_habitat/', x, '.tif'))

  fsize <- file.info(paste0('../results/marginal_habitat/', x, '.tif'))$size/1e9
  if(fsize < 0.025){marg_patch <- terra::patches(marg_patch, directions = 4)} else
    if(fsize > 0.025){marge1 <- terra::patches(terra::aggregate(marg_patch, 2, fun = 'mean'), directions = 4)} else
      if(fsize > 0.04){marge1 <- terra::patches(terra::aggregate(marg_patch, 3, fun = 'mean'), directions = 4)} else
        if(fsize > 0.05){marge1 <- terra::patches(terra::aggregate(marg_patch, 4, fun = 'mean'), directions = 4)} else
          if(fsize > 0.06){marge1 <- terra::patches(terra::aggregate(marg_patch, 5, fun = 'mean'), directions = 4)} else
            if(fsize > 0.07){marge1 <- terra::patches(terra::aggregate(marg_patch, 6, fun = 'mean'), directions = 4)}
  if(exists('marge1')){marg_patch <- terra::resample(marge1, marg_patch, threads = 16, method = 'near')}

  occ_pts <- terra::extract(marg_patch, terra::vect(pts), bind = TRUE) |>
    sf::st_as_sf()
  occ_poly <- dplyr::filter(occ_pts, is.na(patches)) |>
    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
  occ_poly <- terra::extract(marg_patch, terra::vect(occ_poly), weights = TRUE) %>% 
    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
    tidyr::drop_na(patches) |> # these values actually NA. 
    dplyr::slice_max(weight, with_ties = F)

  occ_patches <- c(
    dplyr::filter(occ_pts, ! is.na(patches)) |>
      dplyr::pull(patches), 
    occ_poly$patches
  )

  if(length(occ_patches) > 0){

    marg_patch <- terra::subst(marg_patch, occ_patches,  occ_patches, others = NA)
    old_ID <- unique(marg_patch)$patches # gather and relabel patches 
    lkp <- data.frame(
      old_ID = old_ID,
      new_ID = highest_no + 1:length(old_ID)
    )
    marg_patch <- terra::subst(marg_patch, lkp$old_ID, lkp$new_ID)
    final_patches <- terra::cover(top_patches, marg_patch)

    terra::writeRaster(final_patches, paste0('../results/allOccupiedPatches/', x, '.tif'))
  } else {
    terra::writeRaster(top_patches, paste0('../results/allOccupiedPatches/', x, '.tif'))
  }
}

# lentiformis  NTIF

lapply(f[c(288:335)], combine_pop_patches) # 2:46
```

```{r create a polygon dataset, eval = F}

  f <- paste0('../results/allOccupiedPatches/', 
            list.files('../results/allOccupiedPatches/'))

tif2shp <- function(x){

  taxon <- gsub('.tif$', '',  basename(x))
  pg <- terra::as.polygons(rast(x)) %>% 
    sf::st_as_sf() %>% 
    sf::st_make_valid()

  sf::st_write(pg, paste0('../results/allOccupiedPatchesVector/', taxon, '.shp'), quiet = TRUE)

@@

lapply(f, tif2shp)
